---
title: "tidyrstats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidyrstats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

setwd(here::here())

#devtools::install_github('bansell/tidyrstats')
devtools::load_all()
```

## Description 

This package contains helper functions for implementing common statistical functions in R, within tidy workflows.

## Install

```{r, eval=FALSE}
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}
devtools::install_github("bansell/tidyrstats")
```

```{r setup, eval=FALSE}
library(tidyrstats)
```

```{r,echo=FALSE,message=FALSE}
library(ggplot2)
library(dplyr)
```

## neg_log()

```{r,echo=FALSE,eval=TRUE}
toptags <- readRDS(glue::glue('{here::here()}/local/toptags_slim.Rds'))
```

This function computes the negative logarithm of a numeric input using base 10 by default. This function returns the negative logarithm of `x`. By default, it uses base 10, but you can specify a different base using the `base` argument. Designed for quickly transforming p values for statistical analysis.

Say we have summary statistics results for an RNAseq differential expression experiment

```{r}
head(toptags)
```

We can -log10 transform the P.Value column

```{r}
toptags |> 
  # -log10 transform the P.Value column
  mutate(neg_logP = neg_log(P.Value)) 

```

...then create a volcano plot:

```{r, fig.width = 6,fig.height=4}

toptags |> 
  # -log10 transform the P.Value column
  mutate(neg_logP = neg_log(P.Value)) |> 
  # volcano plot
  ggplot(aes(x=logFC,y=neg_logP)) + 
  geom_point(size=0.5)
```

## scale_this()

This function scales and centres a numeric vector by subtracting the mean and dividing by the standard deviation.

Unlike `base::scale()`, it returns a numeric vector, not a matrix, making it play well with tidy data structures and dplyr chained code. Note this function does not allow control over centering or scaling at present.

`` scale_this` `` will drop NA values before scaling, by default (as is the case for `base::scale()`). However the function will include original NA values in the output, and generate a warning message.

```{r}

iris_dat <- c(head(iris$Sepal.Length),NA)

iris_dat |>  as_tibble() |> mutate(scaled = scale_this(value))


```

`scale_this()` will also work with a simple numeric vector:

```{r}
scale_this(iris_dat)
```

## lm_test()

This function provides a wrapper around `do(broom::tidy(lm(... , data = .))` which is very difficult to teach in a tidyverse-focused workshop; difficult for students to understand, as well as annoying to code!

`lm_test()` is compatible with grouped or nested input, and compatible with the native R pipe `|>` or magrittr pipe `%>%`. It accepts a quoted or unquoted model formula.

`lm_test()` returns a tibble containing the tidied output of `lm()`, sorted by p-value, which includes the following columns:

-   **term**: The name of the model term (such as the intercept, predictors, or interaction terms).

-   **estimate**: The estimated coefficient (also known as the beta value).

-   **std.error**: The standard error associated with the estimate.

-   **statistic**: The t-statistic used in hypothesis testing.

-   **p.value**: The p-value indicating the significance of the term.

If the input data is grouped or nested, the output will retain the group identifiers. In both cases, the grouping/nesting identifiers will appear in the left-most columns of the tibble.

For ungrouped input:

```{r}

mpg %>% lm_test( cty ~ hwy )

```

For grouped input:

```{r}

mpg %>% group_by(class) %>%  lm_test(cty ~ hwy * cyl + 0)

```

For nested input with a quoted model formula:

```{r}
mpg |>  nest_by(class) |>  lm_test("cty ~ hwy * cyl + 0")
```

